Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-06-07 man/ecosystems/ms_ecosystems.Rmd
update intro

## Introduction
The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016]. It has been assumed that geographical and ecological rear-edge populations match. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@VilaCabrera2019; @Pironon2016]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014]. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine under what environmental conditions the geographical rear edge does not correspond with the ecological one [@VilaCabreraJump2019GreaterGrowth].

Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of more intense and longer droughts [@Allen2010]. In fact, drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate-change projections for the Mediterranean area forecast more frequent and severe extreme climate events [@Spinoni2017a]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005]. In a global-change context, the impact of drought on vegetation needs to be evaluated regarding the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape alteration, such as the Mediterranean region, where land-use change plays a major role in current ecosystem ecology [@Navarro2013], given that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].

The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution. In these cases, climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005]. The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010]. Also, it is crucial to accurately characterize the ecological rear-edge of species. This can be particularly misleading in ecosystems where land-use legacies have strongly modified species distribution [@VilaCabrera2019]. Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].

The assessment of resilience to climate disturbances such as droughts provides critical information concerning the capacity of the forests to maintain their structure and render valuable ecosystem services. The drought resilience of species is better assessed by a combination of several approaches, such as dendrochronology and remote sensing. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and annual tree-ring widths can be used to study growth changes in response to drought among individiual trees [@Fritts1976]. Remote sensing can be used to analyze the impact of drought on ecosystems and hence on the stand [*e.g.* @Zhang2013]. The combination of both approaches (remote sensing and dendroecology) has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018]. Nonetheless, it is crucial to ascertain whether the responses at the tree level differ from those at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd. (Pyrenean oak) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades). In the study area, the rear edge of relict populations is imposed by historical land use. Our main hypothesis is that, limited by their land-use legacy, these stands will show low resilience to extreme drought from climate change a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Quercus pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.

@@

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability according to elevation. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002]. Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest forest formation of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].

>>>>

## Introduction

The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016]. It has been often assumed that geographically marginal populations represent ecological marginal populations. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pironon2016; @VilaCabrera2019; @Oldfatheretal2020RangeEdges]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014]. This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and an genetic components [@VilaCabrera2019], but also an anthropogenic (*i.e.* land-use) dimension. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to determine the factors that cause mismatches between the geographical and the ecological rear-edge [@VilaCabreraJump2019GreaterGrowth].

Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sexton2009]. Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allen2010]. Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005]. The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@Hampe2005].

It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Sala2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @Navarro2013; @DoblasMiranda2017]. In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects]. The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLandUse]. For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests]. Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate‐driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth]. In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forest at the southern geographical range. Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLandUse], and particularly that of marginal rear edge populations [@VilaCabrera2019].

The assessment of resilience to climate and human disturbances provides critical information concerning the capacity of forests to maintain their structure and render valuable ecosystem services. Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYoua]. @Lloret2011 proposed an approach, which decomposes resilience to drought in three components: resistance to drought, recovery after drought and resilience. Forest resilience is determined by the capacity to reduce the impact (resistance) and the capacity to recover from the impact of disturbance (recovery) [@IngrischBahn2018ComparableQuantification]. This conceptual approach has been very popular to assess forest resilience recently, because it allows a simple, yet highly efficient assessment of short-term responses of trees to drought. Nevertheless, it is not exempted from some criticism and needs to be applied carefully to avoid over interpretation and potential bias associated at different levels [@Schwarzetal2020QuantifyingGrowth]. In this sense we assessed forest resilience both in the long-term to several recent extreme drought episodes and to climate change (*i.e.* warming) in the long-term (the last few decades) using two different proxies to characterize resilience. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and, generally annual tree-ring widths have been used to study growth changes in response to drought at the individual tree level [@Fritts1976; @Dobbertin2005TreeGrowth]. Remote sensing can be used to analyze the impact of drought on ecosystems and hence at the stand level [*e.g.* @Zhang2013]. The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].

In the present study, we assessed resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be severely threatened by climate change. We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests. To do so we conducted a review of historical documents to reconstruct the forest history of our study sites, and to determine how anthropogenic drivers have shaped the current forest structure. Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local‐scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations. Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter task, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.

@@

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability according to elevation. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002]. Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest forest formation of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG]. They also harbor high levels of intraspecific genetic diversity [@Valbuena2013]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].
